Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-15 appendix.tex
counter-Qs

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The previous upper bound is for the ancestral state split $\{1\}$.
For example, for the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

@@

@@

%EM Shouldn't this be 2p_{\emptyset}-1 ? It doesn't matter for the bound, but...
%das If we call p_{\emptyset} just p, we're maximizing
%   p log (1 + w) + (1 - p) log (1 - w).
%As before, a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
%So then w = (p - (1 - p)) / (p + (1 - p)) = (2p - 1) / 1 = 2p - 1.
%So you're right. Good catch.
%
%Unfortunately, I realized the above does not work for us since 2p - 1 is not guaranteed to be between 0 and 1, as x should be (see the stuff I added justifying this above).
%Rather than do a constrained optimization, I just bounded the p_{\emptyset}\log(1+w) term below by p_{\emptyset}\log(1-w).
%das: This is old and I believe the text has been updated to fix it. Can you verify?

>>>>

Let
$$
%EM It seems a little strange that eta appears here as an argument when it didn't appear in eq:site_pattern_likelihood. In fact it's not really an argument, right? Everything is determined by t and tau it seems to me.
%das: We added it to clarify that the likelihood now depends on \xi, which is an element of \eta. It seems dirty to ignore it. At the risk of adding more notation, we could add \hat{H} to the original likelihood? Or just axe the etas...
%EM I think I must be missing something:
%EM The likelihood doesn't belong on the specific choice of \xi. Besides, \xi already appears in eq:site_pattern_likelihood.
%EM Eta is a deterministic function of t and tau, so isn't having it redundant?
%das: The likelihood doesn't depend on the specific choice of \xi, but it does depend on \eta. I don't know, I guess redundancy isn't always the worst thing? I just see it as a tradeoff between how explicit we want to be and how clean. Like, in the first line of eq:site_pattern_likelihood we could have as an argument \hat{H} instead of suppressing it, even though it's totally determined by the inputs as-is. Anyway, that was my justification for including \eta. I'm fine with removing it though if you think it's weird.
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t)) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}

@@

%EM I think this following statement could use a little more justification. If we write out a symmetric function f in two variables and take the derivative of f(x+delta, x) = f(x, x+delta) with respect to delta and set it to zero we can see that this must be a stationary point of f. Do we need to say more to ensure that this is a maximum?
%das: I wasn't even thinking derivatives. Let f(x, y) be symmetric so that f(x, y) = f(y, x). If f is maximized at some (x_0, y_0) then f(x_0, y_0) > f(x, y) for all x, y. But then by symmetry f(y_0, x_0) is also a maximum. Now *if* a unique maximum exists, x_0 = y_0 should necessarily be this maximum by the properties defined above, otherwise, f(x_0, y_0) would not be equal to f(y_0, x_0). I think there are probably counterexamples where there is not a unique maximum where this statement isn't necessarily true (trivially, because you'd just say x_0 = y_1 if y_1 is also a maximum and voila) but I don't think this discounts the fact that x_0 = y_0 will still be *a* maximum in this case. In that case, it's reasonable to assume so, in my opinion, especially for the topological inconsistency (we don't use the branch length values), but even for the branch length inconsistency because the case of no unique solution is just as bad if not worse than inconsistency.
%das: Should some or all of that be in the text? Can you see anything wrong with it?
%EM I don't see why this works. In fact, imagine that x=y is a ridge of minima, and that we have maxima at (1,2) and at (2,1). My argument shows that we must have a stationary point at x=y, but that doesn't mean that it has to be a maximum!
%das: Wouldn't the assumption of there existing a unique maximum preclude your case? My argument is that if there is a unique maximum, it necessarily has to have x=y. If there is not a unique maximum then the ML problem is ill-posed. I now don't see why we have to have a unique maximum other than to say we hope the problem is well-posed, but it may not be. Do you have any thoughts on where to take this?
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$, $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\label{eq:farris-upper-bound}

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
%das: not quite sure what this comment means actually.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t; \fullAncestralSplitPartitions(\tau_1,t)) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

@@

%EM No, the lower bound is (1+x)(1-x)(1-w)(1+y)(1-y).
%das: I agree that that's *a* lower bound, yeah. So maybe tell me what's wrong with the following: L is the general likelihood in the joint inf problem for the Felsenstein topology, obtained by doing the usual maximization over ancestral states in tab:likelihoods. Now let's obtain L' via the same process but we've now replaced all (1+w)s in that table with (1-w)s. L \ge L', so then L' is a lower bound for the likelihood L, yes? So looking at, let's call it the lower bound version of tab:likelihoods where everything is a (1-w), maximizing over ancestral states then results in the above inequality. Maybe I'm being careless?

@@
